\documentclass[twoside, 12pt]{iiser-thesis}

%Bio dept wants all font in Arial (ew)
%Will keep it commented until the v end for my own sanity. I refuse to actively change a tex file to Arial permanently. Change compiler to XeLaTeX or LuaLaTeX for this to work
%\usepackage{fontspec}
%\setmainfont{Arial}

%my own list of packages
\input{head}

%this file contains the bibliography
\addbibresource{refs.bib}

%%%%%%%%%%%%%%%%%%%
% Packages/Macros %
%%%%%%%%%%%%%%%%%%%
\usepackage{fullpage}
\setlength{\parskip}{1em}

\usepackage{setspace}
\onehalfspacing

\setcounter{tocdepth}{3}

\catcode`\@=11
\catcode`\@=12

%%%%%%%%%%%%
% Document %
%%%%%%%%%%%%
%\setcounter{chapter}{-1}

\title{Eco-evolutionary dynamics of finite populations from first principles}


\author{Ananda Shikhara Bhat}
\coordinator{Coordinator} \supervisor{Vishwesha Guttal}\cosupervisor{Rohini Balakrishnan} \sdesignation{}
\department{Centre for Ecological Sciences, Indian Institute of Science} \reader{Sutirth Dey}
%\reader{Reader 2}
\dedication{This thesis is dedicated to ?}
\graduationyear{2023}
\academicyear{2022-2023}
\graduationmonth{April}

\thesisabstract{
Some central questions of ecology and evolution ask how many distinct variants of an entity can emerge and coexist in sympatry. Examples include the study of standing genetic variation (alleles), polymorphisms (genotypes/phenotypes), and sympatric speciation (species). Historically, these questions have often been studied through phenomenological models formulated for infinitely large populations. Stochastic individual-based models (IbMs), where mechanistic rules are specified for each individual, are more realistic but are rarely studied analytically.\\
\\
In this thesis, I show how ideas from statistical physics can be used to analytically formulate and study IbMs from biological first principles. Starting from a stochastic ‘birth-death process’, I show how density dependence of the population dynamics can be used to obtain a continuous approximation of the system as a stochastic differential equation (SDE). I show that well-known ‘fundamental theorems’ of evolution, such as the Price equation, are recovered in the infinite population limit. My stochastic formulation reveals a new directional evolutionary force in finite populations that lends itself to a simple biological interpretation and has no analogue in infinitely large populations. As an application, I show how this formalism can be used to study cooperation in finite populations. Finally, I extend this theory to study quantitative traits through a stochastic field equation, which can be thought of as an infinite-dimensional SDE. In the infinite population limit, I recover well-known frameworks such as Kimura's continuum-of-alleles and adaptive dynamics. My work thus highlights the connections between various equations in ecology and evolution by revealing their common stochastic underpinnings.
}

\acknowledgments{Not more than 250 words}
\allowdisplaybreaks %allow align envts to go across pages   

\begin{document}
\thesisfront
\listoftables
\listoffigures	
\chapter{Introduction}
\input{ind_files/01_Introduction}
\chapter{Adaptive diversification: a crash course}\label{chap_AD}
\input{ind_files/02_Foundations}
\chapter{Population dynamics from stochastic first principles}\label{chap_BD}
\input{ind_files/03_BD_processes.tex}
\chapter{Examples and Applications}\label{chap_examples}
\input{ind_files/04_examples}
% \chapter{}
% \input{ind_files/chapter_05}
% \chapter{} <-- uncomment and add as required
% \input{ind_files/chapter_06}

\begin{appendices}
\chapter{Deriving the Fokker-Planck equations for It\^{o} SDEs}\label{App_SDE_FPE}
\input{ind_files/99_Appendix_A}
\chapter{Deriving stochastic trait frequency dynamics using It\^{o}'s formula}\label{App_density_to_freq}
\input{ind_files/99_Appendix_B}
\end{appendices}

\printbibliography[title=References]
\end{document}
